Sequence assembly

### 2024-04-08 episodes/phylogenetics.Rmd
A few corrections to msa and phylo. Complete version of assembly/annotation

To perform NJ on our sequences, we'll use the function inbuilt in Seaview. First (re-)open the alignment we obtained from `mafft` with the E-INS-i method, trimmed with ClipKIT.

@@

@@

>>>>

To perform NJ on our sequences, we'll use the function inbuilt in Seaview. First (re-)open the alignment we obtained from `mafft` with the E-INS-i method, trimmed with ClipKIT. If it is not on your computer anymore, transfer it again from Uppmax using `scp` or `SFTP`.

If [Seaview](https://doua.prabi.fr/software/seaview) is not installed on your computer, download it and install it on your computer.

@@

On a Linux computer, you can run it on the command line:

@@

On other computer, just open the file with the regular File > Open menu.

### 2024-04-08 episodes/sequence-assembly.Rmd
A few corrections to msa and phylo. Complete version of assembly/annotation

## Sequence assembly

The assembler we will run is [SKESA](https://github.com/ncbi/SKESA). SKESA generates a final assembly from multiple kmers. A list of kmers is automatically selected by SKESA using the maximum read length of the input data, and each individual kmer contributes to the final assembly. If desired, the lengths of kmers can be specified with the --kmer and --steps flags which will adjust automatic kmer selection.

## Assembly

Because assembly of each genome might take a couple of hours, we will run all assemblies in a loop overnight. It is important to run them within the screen session or else the process will be terminated if we disconnect from the machine.

Preparation

```bash
mkdir ~/molepi/results/assembly
cd ~/molepi/data/trimmed_fastq
```

To run SKESA we will use the skesa command with a number of option that we will explore later on. We can start the loop (again, a short loop) with the assemblies with:

```bash
for sample in ERR029206
do
skesa \
--cores 2 \
--memory 8 \
--fastq "${sample}"_1.fastq.gz_trim.fastq "${sample}"_2.fastq.gz_trim.fastq 1> ../../results/assembly/"${sample}".fasta
done
```

@@

The assembly should take about 2-3 minutes per genome, using 2 cores and 8 GB of memory.

:::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::: instructor

## Downloading the missing assemblies

@@

```bash
cd ~/molepi/results
mv assembly partial_assembly
wget https://github.com/lguy/course-microbial-genomics/raw/main/episodes/files/assemblies.tar.gz .
tar xvzf assemblies.tar.gz
ls assembly
```

@@

:::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::

@@

@@

```bash
assembly-stats ERR026473.fasta
```

```output
stats for ERR029206.fasta
sum = 4233864, n = 451, ave = 9387.73, largest = 70968
N50 = 20633, n = 68
N60 = 16723, n = 91
N70 = 13473, n = 120
N80 = 9592, n = 157
N90 = 5883, n = 213
N100 = 200, n = 451
N_count = 0
Gaps = 0
```

@@

@@

>>>>

## Exercise 1: Sequence assembly with SKESA

The assembler we will run is [SKESA](https://github.com/ncbi/SKESA). SKESA generates a final assembly from multiple kmers. A list of kmers is automatically selected by SKESA using the maximum read length of the input data, and each individual kmer contributes to the final assembly. If desired, the lengths of kmers can be specified with the `--kmer` and `--steps` flags which will adjust automatic kmer selection.

Because assembly of each genome would take a long time, you will only run one assembly here.

### Preparation

Start the interactive session, go to the `molepi` subfolder, create a `results/assembly` subfolder. Then go to the folder containing the fastq files.

```bash
interactive -A uppmax2024-2-10 -M snowy -t 04:00:00
cd /proj/g2020004/nobackup/3MK013/<username>/molepi/
mkdir results/assembly
cd data/trimmed_fastq
```

To run SKESA we will use the skesa command with a number of option that we will explore later on. We can start the loop (again, a short loop with only one sample in this actual case to spare computing resources) with the assemblies with:

```bash
module load bioinfo-tools SKESA
for sample in ERR029206
do
skesa \
--cores 2 \
--memory 8 \
--fastq "${sample}"_1.fastq.gz_trim.fastq "${sample}"_2.fastq.gz_trim.fastq 1> ../../results/assembly/"${sample}".fasta
done
```

@@

The assembly should take about 5 minutes per genome, using 2 cores and 8 GB of memory.

### Downloading the missing assemblies

@@

```bash
cd ../../results/
mv assembly partial_assembly
tar xvzf /proj/g2020004/nobackup/3MK013/data/assemblies.tar.gz
ls assembly
```

@@

@@

### Examine the results of the assembly

You will now explore the files output by SKESA. Find out the number of contigs, and the various assembly statistics.

What do the they mean? What are N50, L50, N80 and L80?

@@

```bash
module load GenomeTools 
gt seqstat ERR026473.fasta
```

```output
# number of contigs:     390
# total contigs length:  4239956
# mean contig size:      10871.68
# contig size first quartile: 989
# median contig size:         5080
# contig size third quartile: 15546
# longest contig:             74938
# shortest contig:            200
# contigs > 500 nt:           323 (82.82 %)
# contigs > 1K nt:            291 (74.62 %)
# contigs > 10K nt:           148 (37.95 %)
# contigs > 100K nt:          0 (0.00 %)
# contigs > 1M nt:            0 (0.00 %)
# N50                25167
# L50                54
# N80                11379
# L80                129
```

@@

## Challenge: Compare the results of the different samples

How could you get N50 stats for all the assemblies?

::: solution

```bash
for accession in ERR029207 ERR029206 ERR026478 ERR026474 ERR026473 ERR026481 ERR026482
do
  echo "${accession}"
  gt seqstat "${accession}".fasta | grep N50
done

```

::::::::::::

:::hint

Hint: https://en.wikipedia.org/wiki/N50,_L50,_and_related_statistics ¨¨

:::::::

@@

## Exercise 2: assembly with SPAdes

To compare how different assemblers perform, you will also use [SPAdes](https://ablab.github.io/spades/). SPAdes is a very versatile and can assemble many different data types, including combining long and short reads.

:::challenge

## Challenge: Assemble the genomes with SPAdes

You will only perform the assembly on one genome, but use the loop form anyway. You will use the `--isolate` mode, but the `--careful` mode would also be interesting. This might take 10-15 minutes per genome.

::: solution

```bash
module load spades
cd /proj/g2020004/nobackup/3MK013/<username>/molepi/results
mkdir spades_assembly
cd spades_assembly
for accession in ERR029206
do
  echo "${accession}"
  spades.py \
    --isolate \
    -1 ../../data/trimmed_fastq/"${accession}"_1.fastq.gz_trim.fastq \
    -2 ../../data/trimmed_fastq/"${accession}"_2.fastq.gz_trim.fastq \
    -o "${accession}"_spades \
    -t 2 \
    -m 8
done

```

::::::::::::

## Challenge: Compare the results of the assembly with SPAdes and SKESA

Using Genome Tools, compare the result of the two assemblies

::: solution

```bash
cd /proj/g2020004/nobackup/3MK013/<username>/molepi/results
gt seqstat assembly/ERR029206.fasta
gt seqstat spades_assembly/ERR029206_spades/contigs.fasta
```

::::::::::::

## Exercise 3: Preliminary annotation

### Physical annotation

The goal here is to find ORFs and start codons, to find proteins in the genome.

::: challenge

## Challenge: Use prodigal

Find how prodigal works and use it to annotate the proteome of ERR029206.

```bash
prodigal -h
```

::: solution

```bash
cd /proj/g2020004/nobackup/3MK013/<username>/molepi/results
mkdir annotation
cd annotation
module load prodigal
prodigal -i ../assembly/ERR029206.fasta -a ERR029206.faa > ERR029206.gbk
```

::::::::::::

::::::::::::

Examine the resulting files, both the `.faa` and the `.gbk`. Can you make sense of the information there?

### Functional annotation

The physical annotation provided by prodigal identifies genes (ORFs and start codons), but it does not give any information about the function of the genes identified. The next step is thus to infer a function for each of the identified genes. This process usually involves aligning the proteins to different databases, e.g. using BLAST and other programs. You will not perform a full functional annotation in this exercise, but you will try to use blastp to find a RpoB (RNA polymerase B) homolog in the genome.

::: challenge

## Create a BLAST database

With the proteome of sample ERR029206, create a BLAST database. Use the `makeblastdb` function that comes with the BLAST package.

::: solution

```bash
module load
makeblastdb -dbtype prot -in ERR029206.faa -out ERR029206
```

::::::::::::

## Find rpoB homologs

Now that you made the database, use the RpoB sequence from *E. coli* that you retrieved in the BLAST exercise to find whether there are one or several RpoB homologs in the proteome you just annotated. Examine the BLAST results: how many homologs to RpoB in that genome?

::: solution

```bash
blastp -query ../../../blast/rpoB_ecoli.fasta -db ERR029206
```

Just one has E-values at the right level.

:::::::::::::

:::::::::::::
